---
title: "Data Organization Basics"
author: "DTK"
date: "CVC Workshop: July 2013"
output: ioslides_presentation
---

```{r include=FALSE}
require(mosaic)
require(DCFdevel)
require(knitr)
opts_chunk$set( tidy=FALSE, prompt=FALSE, comment=NA)
```

## Data Comes in Many Forms

* Images: Still and Video
* Free text: e.g. books, college presidents' speeches, twitter feeds, ...
* Rectangular arrays

### We will work only with rectangular arrays

Call these: **data tables**.  

You'll also hear about one R representation of data tables: **dataframes**.

## Enforce a Strict Convention

Horizontal | Vertical 
-----------|-----------
**Cases**  | **Variables**
Rows       | Columns
Instances  | Fields
Tuples     | Records

## Example: ZIP Codes

```{r include=FALSE}
set.seed(100)
```

```{r}
data(ZipGeography)
ZipGeography %>% 
  select(ZIP, State, Population, LandArea, WaterArea) %>% 
  sample_n(size = 8) %>%
  print( row.names=FALSE)
```

## Big and Small

Some data tables have many cases, some have many variables, some both, some neither.

Examples: 
```{r}
data(ZipDemography)
nrow(ZipDemography)
ncol(ZipDemography)
```

***

```{r}
data(registeredVoters)
nrow(registeredVoters)
ncol(registeredVoters)
```


## Variable Types

* Quantitative: Numeric
    * Just the number, no units
    * No formatting, e.g. ~~~ $1,432,654 ~~~
    * Something you can think of as continuous.
    * Something where the idea of a mean makes sense.
* Categorical: Names or categories
    * Typically a few categories
    * No natural order, e.g. Danish is not between Chinese and English
    * Easy to mis-spell, so cleaning is often necessary.
* Ordinal: Like categorical but with a natural order.
    * Example: Likert Scale   
Hate:::Dislike:::Neutral:::Like:::Love

## Codebook

All the information about units, explanations, etc. go in another file: the codebook.

## Do's & Don'ts

* Do use mnemonic levels for categorical variables:
   * e.g. `F` and `M`
   * e.g. `Never Smoked`, `Past Smoker`, `Curr Smoker`
   
* Don't copy over data from other tables 
   * e.g. I've got the species and want to add the typical weight.
   
* Do use proper data manipulation commands to add data from other tables --- do this when you need it, not before.

* Do consider using multiple tables with different sorts of information.

* Do provide potentially useful auxilliary variables:
    * Who handled this sample
    * The date it was handled
    * Page from the lab notebook
    * any covariates of potential interest, even if you don't plan right now to use them
    
## Cases

For any data table, you should know what the case is:

### Examples

* A voter
* A zip code
* A student in a specific course
* A hospital providing a specific DRG

## Wide and Narrow form

```{r}
data(MigrationFlows)
MigrationFlows %>% select(sex,origincode,destcode,Y2000) %>%
  sample_n(size=10) %>% 
  print(., row.names=FALSE)
```


## Design your Data to be Extensible

Example: Repeated measurements from rats

Animal | Run 1 | Run 2 | Run 3
-------|-------|-------|------
D67    | 17    | 19    |
N01    | 23    | 14    | 11
M03    | 6     |       |
D67    | 15    | 12    |

Questions: 

* What if you wanted to add in a date of the run?
* How would you handle covariates? (e.g. the person conducting the run)

## The Same Measurements: But Extensible 

Animal | Run | Speed  | Technician | Date       | Temperature
-------|-----|--------|------------|------------|-------------
D67    | 1   | 17     | Paul O.    | 2014-07-04 | 31
D67    | 2   | 19     | Steve S.   | 2014-07-06 | 34
N01    | 1   | 23     | Steve S.   | 2014-05-03 | 28
N01    | 2   | 14     | Steve S.   | 2014-05-06 | 30
N01    | 3   | 11     | Paul O.    | 2014-05-19 | 31
...    | ... | ...    | ...        | ...        | ...

and so on.

## Activities

Links on the CVC [Workshop Schedule](http://dtkaplan.github.io/CVC/Summer2014/schedule2014.html).

1. Critique a [file's organization](http://vote.minneapolismn.gov/www/groups/public/@clerk/documents/webcontent/wcms1p-126724.xlsx)
2. More complicated situation. [Military Families](https://dl.dropboxusercontent.com/u/5098197/Math155/ActiveDuty_MaritalStatus.xls)
#. Enter some data

```{r}
sheet1 <- fetchGoogle("https://docs.google.com/spreadsheet/pub?key=0Am13enSalO74dFRFZHNlaVE3NjZfUjBQOUNoMmRlUWc&single=true&gid=0&output=csv")
sheet2 <- fetchGoogle("https://docs.google.com/spreadsheet/pub?key=0Am13enSalO74dFRFZHNlaVE3NjZfUjBQOUNoMmRlUWc&single=true&gid=1&output=csv")
```