---
title: "Getting Data Into R"
date: "CVC 2015"
author: R Pruim
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, include = TRUE, echo = TRUE)
require(mosaic)
trellis.par.set(theme = col.mosaic())
require(tidyr)
```


Working with data in packages is great for learning how R works, but 
eventually you will want to use your own data, or data that you obtain 
from somewhere else.  These notes discuss ways to get data into R.

# Some sources for data

Here are some places that have data sets available online

  * [data.gov](https://catalog.data.gov/dataset?res_format=CSV) has more than 
  10,000 data sets in csv format (plus others in other formats)
  
  * [publicdata.eu](https://publicdata.eu/dataset) has more than 
  10,000 data sets in csv format (plus others in other formats)
  
  * [World Bank](http://data.worldbank.org/) has data in several
  formats, including csv an Excel
  
  * [datahub](http://datahub.io/dataset?res_format=CSV) has over 
  1,000 data sets in csv format (plus others in other formats)
  
  * [CSV format data from several R packages](https://vincentarelbundock.github.io/Rdatasets/datasets.html) 
    * It is easier to get this data from within the package, but 
    they can be used as examples for import from CSV format.
    The resulting data is typically all tidied and ready to be 
    used for analysis.
  
## Importing CSV Data

Data can be stored in many file formats.  The CSV (comma separated values)
and TSV (tab separated values) file formats are among the most commonly used,
and most data handling software can export to CSV and import from CSV, so it 
good to learn how to work with CSV data.

**Exercise:** Obtain a csv file containing data.  You can create
your own csv file from a google or Excel spreadsheet, or you can find a 
CSV file online.

Note: 
If you have data in an Excel spreadsheet, you can "save as" csv.  This isn't
exactly the same thing as exporting as csv because the open file switches
formats and you are no longer working with a regular Excel file.

### Uploading Files to RStudio Server

*This step is only required in the web version of RStudio when the 
file you want to import is on your local machine and not on the server.

 1. In the Files tab, navigate to the folder where you want to store the file.
 2. Choose "Upload File"
 3. Navigate to the file on your local machine.
 
That's all there is to it.  You can use this procedure to upload files of any type.

### Importing Data into your R Session

 1. In the Environment Tab find "Import Dataset".  
 2. Choose "From Text File" or "From Web URL"
 
    * Note: text files must live on the same machine as RStudio.  If you
    are using the web version of RStudio, you may need to upload your file
    first.  See above.
    
 3. Navigate to the file, or paste in the URL.
 4. Name the data set (the default name is derived from the file name).
 5. Fill out the form to declare the structure of the file.
 6. Hit "Import" to import the data.

This will place the data set in your working environment.

### Using R commands directly

Using RStudio's ability to import data is handy for interactive use,
but it is not part of a reproducible workflow an can't be use in
RMarkdown documents.  So it is good to learn how to use R commands 
to import CSV files.

*Note: When you use RStudio's interactive interface, the command used to read the file is printe in the console, so you can also import interactively and copy this command into your RMarkdown document.*

The main functions we can use for this are `read.csv()` 
and `readr::read_csv()`.  These functions work similarly.

For our example, we will use the `Anscombe.csv` file which you can obtain using the command below.

```{r, eval=FALSE}
download.file(
  "http://vincentarelbundock.github.io/Rdatasets/csv/car/Anscombe.csv",
  "Anscombe.csv")
```
```{r}
Anscombe1 <- read.csv("Anscombe.csv", header=TRUE)
require(readr)
Anscombe2 <- read_csv("Anscombe.csv")
```

The following shows a few of the differences between the two functions.

```{r}
glimpse(Anscombe1)
glimpse(Anscombe2)
```

  * `read.csv()` needs to be told that there is a header row containing
  the variable names.
  
  * `read.csv()` modifies variable names to make them "nice";  
  `read_csv()` does not.  
  Notice that in `Anscombe2` has no name for the first variable, 
  while `Anscombe1` renames this as `X`.
  
  * `read.csv()` converts character data into factor data (by default);
    `read_csv()` does not.

```{r}
names(Anscombe2)[1] <- "state"
xyplot( education ~ income, data = Anscombe2)

```

## Importing Excel Spreadsheets

It is also possible to import data directly from an Excel spreadsheet without
first saving it as a CSV file.  
This has some important advantages:

  * It avoids having multiple copies of the data, which could become out of sync and cause confusion.
  * It makes it easy to automate work with Excel spreadsheets.
  * It allows you to work with Excel spreasheets even if you don't have 
  Excel running (or even installed) on your machine
  * If we use the `readxl` package for this, then certain oddities of 
  Excel files are taken care of for us.  
  Most importantly, `readxl` takes care of 
  date conversion an knows how to handle data from multiple versions of 
  Excel.
  
```{r, eval=FALSE}
download.file(
  "http://api.worldbank.org/v2/en/topic/17?downloadformat=excel",
  "Gender.xls")
```

The first two rows of this spreadsheet don't contain data, so we'll skip
them.

```{r}
require(readxl)
Gender <- read_excel("Gender.xls", skip=2)
glimpse(Gender[, 1:10])
```

We would like to reformat this data so that year is one of the variables
instead of having a column for each year.  We can do this with
the `tidyr` package; we just need to inicate which columns should be
converte to data an give a name for the variable (`key`) and its 
values (`value`).  The remaining variables are duplicated as needed.

```{r}
require(tidyr)
Gender2 <- 
  Gender %>% 
  gather(
    "year", "value",
    `1960`:`2014`
  ) %>% 
  filter(!is.na(value)) %>%
  mutate(year = as.numeric(as.character(year)))
glimpse(Gender2)
```

Now we can pick one of the inicators and some countries and compare
them over time.

```{r}
Gender2 %>% 
  filter(`Country Code` %in% c("USA", "NLD", "SWE", "KOR", "JPN"),
         `Indicator Code` %in% c("SP.ADO.TFRT")) %>%
  xyplot(value ~ year, data = ., groups = `Country Code`, type = "l",
         auto.key=list(lines=TRUE, points=FALSE, columns=5), 
         ylab = Gender2$`Indicator Name`[1])

Gender2 %>% 
  filter(`Country Code` %in% c("USA", "NLD", "SWE", "KOR", "JPN"),
         `Indicator Code` %in% c("SP.ADO.TFRT")) %>%
  ggplot() +
  geom_line(aes(y=value, x=year, colour=`Country Code`)) +
  labs(y = Gender2$`Indicator Name`[1])
```
  
  **Exercise:** Import an Excel spreadsheet into R an create a plot.

## Importing from Google Spreadsheets

