Gapminder Data
======================

```{r include=FALSE}
require(DCF)
require(rworldmap)
```

```{r echo=FALSE}
worldMap <- function(form,data=NULL,region=c("World","Eurasia","Africa","Latin America", "North America","Oceania","Asia")){
  cname <- all.vars(rhs(form))
  vname <- all.vars(lhs(form)) 
  region <- tolower(match.arg(region))
  s <- joinCountryData2Map(data, joinCode='NAME',nameJoinColumn=cname )
  par(mai=c(0,0,0.2,0), xaxs='i',yaxs='i') 
  mp <- mapCountryData(s, nameColumnToPlot=vname, addLegend=FALSE,mapRegion=region)  
  do.call(addMapLegend, c(mp, legendWidth=.5,legendMar=2))
}
```
The <gapminder.org/data> site provides access to more than 500 country-by-country "indicator" variables.

In the *Data and Computing Fundamentals* course, we use these data for several purposes:
* Illustrate wide versus long format
* Demonstrate grouping operations
* Demonstrate joining operations
* Visualization, especially maps and scatter plots
* Modeling, especially clustering

Eventually, we plan to write interface software to get data directly from the Gapminder site, but for the present we have downloaded a subset of data and provided access through the `fetchData()` and `fetchGapminder()` functions.

The list of available files is [here](http://htmlpreview.github.io/?https://github.com/dtkaplan/DataAndComputingFundamentals/blob/master/Notes/Day3/GapminderFiles.html), along with instructions for adding new files.

### Wide vs Long

The Gapminder data are originally stored with countries being the cases and years the "variables."  For example, here is the data on cell phone subscriptions (only part of the data is shown):
```{r echo=-2}
phones <- fetchData("Gapminder/CellPhoneTotal.csv")
head(phones)[,c(1,46:48)]
```

This is a "wide" format.  In this presentation, there is a separate variable for each year --- 2009, 2010, and 2011 are shown.  Under that year, the value gives the number of cell phone subscriptions.

Although the wide format is reasonable, it has disadvantages.  In particular, it can be hard to do comparisons across countries, except for the same year.

The "narrow" format of these same data has three rows: country, cell phone number, and year.  Converting between wide and narrow formats is not difficult, but to simplify things, we have presented a function that does the conversion automatically:
```{r echo=-2}
phones <- fetchGapminder("Gapminder/CellPhoneTotal.csv")
head(subset(phones, Year >= 2009 & (Country %in% c("Afghanistan","Albania","Algeria","Argentina","Aruba"))),18)
```

To a human reader, this format is not necessarily easier than the wide format. But with grouping and joining operations, it's considerably easier.

For instance, suppose you're interested in the fraction of the population with cell phone subscriptions.  You'll want to read in both the phone and the population data, sorting them out year-by-year for each country.  Then divide subscriptions by population to get an index, and perhaps display this as a map.  But not every country has a reading for every year, so perhaps best to look at the maximum number reported for the period since 2005.

```{r warning=FALSE,message=FALSE}
phones <- fetchGapminder("Gapminder/CellPhoneTotal.csv")
phonemax <- groupBy(phones,by=Country,phones=max(CellPhoneTotal))
pop <- fetchGapminder("Gapminder/TotalPopulation.csv")
# Bad formatting by Gapminder
pop <- transform(pop, pop=as.numeric(gsub(",","",TotalPopulation)))
popmax <- groupBy(pop,by=Country,pop=max(pop,na.rm=TRUE))
both <- join(popmax,phonemax)
both <- subset(both, phones > 0)
both <- transform(both, phoneDensity=phones/pop)

```

Mapping out the number of subscriptions per population shows some expected and some perhaps unexpected patterns, such as the large density in countries such as Chile, Argentina, Libya, South Africa, Zimbabwe, etc.
```{r}
worldMap(phoneDensity ~ Country, data=both)
```


