---
title: "Modeling in R"
author: ""
date: ""
output: 
  html_document:
    fig_height: 3
    fig_width: 5
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
require(mosaic)   # Load additional packages here 
require(ggformula)
theme_set(theme_bw(base_size = 9))

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```


## Modeling in R

### (Simple) Linear Regression

Let's predict length from width.

#### Begin with a plot
```{r}
head(KidsFeet)
gf_point(length ~ width, data = KidsFeet) %>%
  gf_lm(length ~ width)
```

#### Fit a model and summarize

Here is the summary several times over using different styes.  `msummary()` gives a more minimal summary.

```{r}
Kids.model <- lm(length ~ width, data = KidsFeet)
msummary(Kids.model)
Kids.model %>% msummary() 
lm(length ~ width, data = KidsFeet) %>%
  msummary()
```

#### ANOVA table if you like

```{r}
anova(Kids.model)
```

#### Function-based predictions

```{r}
# create a function
predicted.length <- makeFun(Kids.model)
# use it to make a prediction
predicted.length(width = 8.5)
# use it to make multiple predictions
predicted.length(width = c(8, 8.5, 9))
# or predict for all widths in the data set
fitted(Kids.model)
```

#### But wait, there's more -- interval predictions

```{r}
predicted.length( width = 9, interval = "confidence")
predicted.length( width = 9, interval = "prediction")
gf_point(length ~ width, data = KidsFeet) %>%
  gf_lm(length ~ width) %>%
  gf_pointrange(24.73 + 24.4 + 25.06 ~ 9, color = "red")
gf_pointrange()
apropos("gf_")
```
#### Automated intervals for coefficients

```{r}
confint(Kids.model, level = 0.9 )
```

### Adding some predictors

```{r}
Kids.model2 <- lm( length ~ width + domhand, data = KidsFeet)
msummary(Kids.model2)
mplot(Kids.model2, which = 1:7, system = "gg")
gf_lm(length ~ width, color = ~domhand, data = KidsFeet)
plotModel(Kids.model2, system = "gg")
```


### Adding interaction

```{r}
Kids.model3 <- lm( length ~ width * domhand, data = KidsFeet)
msummary(Kids.model3)
plotModel(Kids.model3, system = "gg")
```

Take a look in the `effects` package for a general way to plot complicted models.

### Higher order polynomial terms and transformations

```{r}
Kids.model4 <- lm( log(length) ~ poly(width,2) * domhand, data = KidsFeet)
msummary(Kids.model4)
```

### Generalized linear models

```{r}
glm(sex ~ length * domhand, data = KidsFeet, family = binomial(link = logit)) %>%
  msummary()
glm(sex ~ length * domhand, data = KidsFeet, family = binomial(link = logit)) %>%
  plotModel(system = "gg")
```


### What about ANOVA

```{r}
lm( length ~ sex * domhand, data = KidsFeet) %>%
  anova()
aov( length ~ sex * domhand, data = KidsFeet) 
```

```{r}
xyplot(length ~ sex, groups = domhand, type = c("a", "p"), data = KidsFeet)
gf_point(length ~ sex, color = ~domhand, data = KidsFeet) 
#  gf_line(length ~ sex, color = ~domhand, data = KidsFeet, stat = stat_summary())
```

### A nonlinear model

```{r}
nls( length ~ a + width^p, data = KidsFeet, start = list(a = 0, p = 1)) %>% 
  summary()
```



