---
title: Every Tom, Dick, and Harry
author: Writing Functions in R
date: "CVC 2015"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r, include = FALSE}
require(mosaic)
require(babynames)
```

## What ever happened to Jane (and Mary)?

```{r}
babynames %>% 
  group_by(sex, year) %>% mutate(total = sum(n)) %>%
  group_by(name, sex, year) %>% mutate(prop2 = n / total) %>%
  filter(name %in% c("Mary", "Jane"), sex == "female") %>%
  ggplot() +
  geom_line(aes(x = year, y = prop, color = name))
```

## What ever happened to Jane (and Mary) (and Mary Jane)?

```{r}
BabyNames2 <-
  babynames %>% 
  group_by(sex, year) %>% mutate(total = sum(n)) %>%
  group_by(name, sex, year) %>% mutate(prop2 = n / total) 
```

```{r}
BabyNames2 %>%
  filter(name %in% c("Mary", "Jane"), sex=="F") %>%
  ggplot() +
  geom_line(aes(x = year, y = prop, color = name))
```



## What about Tom, Dick, and Harry?

We would like to generalize our code to handle any set of names
we want to compare.  In R we do this by writing **functions**.

We begin with our two questions:

 1. What do we want R to do?
   
    * This determines the name of the function (and what it does/returns)
    * `NameTrendPlot()` might be a good name for the function
 
 2. What does it need to know to do this?
 
    * This determines the arguemnt list (list of inputs) to the function
    * The function will need to know which names to track.  
    * We should probably have an argument to restrict to just one sex.
    
## Creating a function in R

The general form of a fucntion definition is

```{r}
functionName <- function( arg1 = default1, arg2 = default2, ... ) {
  # do stuff here
  # result of last line is returned.  
  # Can also use explicit return().
}
```

## What about Tom, Dick, and Harry?

```{r}
NameTrend <- function(names, sex) {
  BabyNames2 %>%
    filter(name %in% names, sex %in% sex) %>%
    ggplot() +
    geom_line(aes(x = year, y = prop, color = name, linetype=sex))
}
```

## What about Tom, Dick, and Harry?

``{r}
NameTrend(c("Tom", "Dick", "Harry"), sex=="M")
```



