---
title: "Modeling in R"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "June 23, 2016"
output: pdf_document
---

This document describes ways to fit a variety of models using the `mosaic` package.  See https://github.com/ProjectMOSAIC/LittleBooks/blob/master/README.md for a link to the `Student Guide to R` that provides more details about linear regression modeling.

```{r message=FALSE}
options(digits=3)
require(mosaic)
require(NHANES)
```


```{r}
favstats(~ female, data=HELPrct)
tally(~ sex, data=HELPrct)
tally(~ sex, format="percent", data=HELPrct)

mean(pcs ~ sex, data=HELPrct)
bwplot(pcs ~ sex, data=HELPrct)
```

Now let's fit a multiple regression model for PCS (physical component scores) that includes mcs, sex, and substance (3 levels).
```{r}
mlrmaineffect <- lm(pcs ~ mcs + sex + substance, data=HELPrct)
coef(mlrmaineffect)
msummary(mlrmaineffect)
mplot(mlrmaineffect, which=7)
```

Let's do some model diagnostics.

```{r}
mplot(mlrmaineffect, which=1)
histogram(~ resid(mlrmaineffect), fit='normal')
```

Let's plot some predicted values, say for a alcohol involved subject, as a function
of being male vs. female and MCS score.


```{r fig.keep="last"}
mlrmefun <- makeFun(mlrmaineffect)
xyplot(pcs ~ mcs, data=HELPrct)
plotFun(mlrmefun(mcs, sex="female", substance="alcohol") ~ mcs, add=TRUE, col="black")
plotFun(mlrmefun(mcs, sex="male", substance="alcohol") ~ mcs, add=TRUE, col="red")
```

```{r}
plotModel(mlrmaineffect)
```

But for something even more interesting

```{r}
diabmod1 <- glm(Diabetes ~ Age + BMI, family="binomial", data=NHANES)
diabfun1 <- makeFun(diabmod1)
plotFun(diabfun1(Age=Age, BMI=BMI) ~ Age + BMI, xlim=c(0, 80), ylim=c(15, 40))
```

```{r}
diabmod2 <- glm(Diabetes ~ Age + BMI + Age*BMI, family="binomial", 
  data=NHANES)
diabfun2 <- makeFun(diabmod2)
plotFun(diabfun2(Age=Age, BMI=BMI) ~ Age + BMI, xlim=c(0, 80), ylim=c(15, 40),
        main="Difference in predicted probabilities of diabetes (Interaction Model)")
```


Contour plots...

```{r}
plotFun((diabfun2(Age=Age, BMI=BMI) - diabfun1(Age=Age, BMI=BMI)) ~ Age + BMI, xlim=c(0, 80), ylim=c(15, 40), main="Difference in predicted probabilities\nof diabetes (Interaction - Main Effect)")
```




