---
title: "Scraping Data"
author: "R Pruim"
date: "July 10, 2014"
output: html_document
---

## CIA World Factbook

Google for [CIA factbook](https://www.cia.gov/library/publications/the-world-factbook/)a

*  Go to references and then Guide to Country Comparisons and you will 
land [here](https://www.cia.gov/library/publications/the-world-factbook/rankorder/rankorderguide.html)
    * there will be lots of data sets you can navigate to
    * each one will have a download link to get to the data, 
    for example at [https://www.cia.gov/library/publications/the-world-factbook/rankorder/rawdata_2186.txt](https://www.cia.gov/library/publications/the-world-factbook/rankorder/rawdata_2186.txt)

We can read this with
```{r}
require(RCurl)
url <- "https://www.cia.gov/library/publications/the-world-factbook/rankorder/rawdata_2186.txt"
ciaData <- read.delim(textConnection(getURL(url)))
head(ciaData)
```

### Wikipedia FIFA world rankings

[http://en.wikipedia.org/wiki/FIFA_World_Rankings](http://en.wikipedia.org/wiki/FIFA_World_Rankings) has some tables that we would like to scrape.

We need the `XML` pacakge for this
```{r}
require(XML)
```

```{r}
url <- "http://en.wikipedia.org/wiki/FIFA_World_Rankings"
fifa <- readHTMLTable(url, stringsAsFactors=FALSE)
str(fifa, max.level=1)
length(fifa)
```

### Brainy Zip

```{r}
url <-  "http://www.brainyzip.com/zipcodes/53/53001.html"
zip <- readHTMLTable(url)
str(zip, max.level=1)
zip[[2]]
```
