---
title: "CVC 2014 Schedule"
date: "July, 2014"
output: html_document
---


```{r include=FALSE}
options(markdown.HTML.stylesheet = '../stylesheets/cvc2013.css')
options(rstudio.markdownToHTML =
  function(inputFile, outputFile) {     
    require(markdown)
    markdownToHTML(inputFile, outputFile, stylesheet='../stylesheets/cvc2013.css')  
  }
)
```
```{r eval=FALSE, echo=FALSE}
<script src="https://ajax.googleapis.com/ajax/libs/jquery/1.7.1/jquery.min.js"></script>
<script src="../javascripts/main.js"></script>
```


Working Schedule, July 7-11, 2014 
========================================================

### Resources
* `mosaic` package (version 0.9-0 or greater) and its vignettes are [available via CRAN](http://cran.r-project.org/web/packages/mosaic/index.html) and also via [github](https://github.com/rpruim/mosaic)
* [The Calvin College RStudio server](http://acc.calvin.edu) has provided accounts for 
all workshop participants.
* Data and Computing Fundamentals [spring 2013 course syllabus](http://htmlpreview.github.io/?https://github.com/dtkaplan/DataAndComputingFundamentals/blob/master/Spring2013/syllabus2013.html)
* Materials for our [USCOTS 2013 Workshop](http://uscots2013.mosaic-web.org/Workshop) on
Teaching Statistics Using R.
* Slides and files for sessions will be posted in the schedule as we go along
* Videos
    * Vine video about graphics, like [this one](https://vine.co/v/MI5DM173wWz) and [this one](https://vine.co/v/MI5EWixM65A) about glyphs.
* [ggplot2 documentation](http://docs.ggplot2.org/current/) shows 
examples of each geom, stat, scale, etc.

### Provide Feedback
* Give a suggestion/correction: [Google Doc](https://docs.google.com/document/d/1eEwaMvxahJJ2YuFFWgUEmgBj58umwvMwf_GXnsp9Yc0/edit?usp=sharing)  
* Share/send some code via Firepad:  [Alpha](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-alpharIwB7uaxo82e5QET), [Bravo](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-bravoIwB7uaxo82e5QET), [Charlie](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-charlieIwB7uaxo82e5QET), [Delta](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-deltaIwB7uaxo82e5QET), [Echo](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-echoIwB7uaxo82e5QET), [Foxtrot](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-foxtrotIwB7uaxo82e5QET), [Golf](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-golfIwB7uaxo82e5QET), [Hotel](http://www.mosaic-web.org/go/firepad/examples/codeCOMP121.html#-hotelIwB7uaxo82e5QET).  


--------------------------

## Sunday

Arrival and opening dinner

## Monday

1. Welcome and Introductions
    a. Paul Overvourde
    b. goals 
        * By Wednesday noon you will have created a document containing
            * text
            * at least one plot
            * a description of something interesting to you
            * R will let you create this as a PDF, HTML, or Word file. Also
            options to produce slides insted of a document.

        * By Friday you will have created at least one document 
        using R to explore data
            that can be used in your course, lab, research, etc.
            * may be expansion of Wednesday's or completely different
            * may be multiple little things or one larger thing
            
    #. outline of the week [who?]
        * first 2.5 days -- mostly shop, a bit of work
            * we have prepared some things that we think 
            will be generally useful
            * basic R/RStudio
            * data manipulation
            * data visualiztion
            
        * second 2.5 days -- mostly work, a bit of shop
            * we will work with you to find and learn what you need
     

#. A First Case Study -- CIA case book [Danny]
    a. Randy suggests that this be R free -- focus on tasks and products but don't show any code (this is to motivate learning some R skills to come)
    b. Important ingredients
        * data scraping
        * graphical displays
        * modeling
        * other?
  
#. RStudio orientation 1 [Randy (unless Nick is there already)]
    a. logging in; changing password
    #. arithmetic and syntax
    #. getting help
    #. panes and tabs

#. [Less Volume, More Creativity](Slides/LessVolume.html) -- An R Quick Start [Randy]
    a. Principles for successfully using R with students
        * Less Volume, More Creativity 
        * Two Questions
    
    #. The formula-interface template
        * graphical summaries with `lattice`
        * numerical summaries with `mosaic`
        * foreshadowing modeling with `lm()` and its cousins
        
    #. slides include stuff that will probably not be covered there
        * linear models
        * eXtras like `xpnorm()`, `xchisq.test()`
        * resampling with `do()`

#. RMarkdown [Nick]
    a. Creating PDF/HTML/Word documents in RStudio
    #. **Activity**: Write and publish a description of some straightforward, data-oriented topic you'd like to present in class.  Keep it simple.  We'll do more complex things in the 2nd half of the workshop.
        * include at least one plot  -- but it may be unrelated
    
#. Basic R building blocks 
    a. basic data types: numeric, character, factor, logical
    #. some special objects: `NA`,  `NaN`, and `NULL`
    #. containers
        * vectors, lists, and data frames
    #. objects and classes
    #. names
        * variable names versus strings
    
#. Data Organization
    a. rectangular data: case=row, variable=column
        * imagine a spreadsheet-like table, but
        * Note: many excel spreadsheets are **not** rectangular data
    #.  **Activity**: The spreadsheet [here](http://vote.minneapolismn.gov/www/groups/public/@clerk/documents/webcontent/wcms1p-126724.xlsx) contains data on the Minneapolis 2013 election by ward and precinct.  Identify the elements of this spreadsheet that are not in standard form for a data tabl
    #. will say more about getting data into R tomorrow
    #. wide vs narrow formats -- what is a case?
        * do we want to talk about using R to convert here or just 
            about the alternative formats?  (This could be discussed
            in the context of how data are organized in Excel.)
        * NCI60 [to complicated for here?] and Aaron's wind matrices
        
#. The graphics-data interface -- graphing concepts
    a. goal: more flexible creation of customized plots
    
    #. parts of graphics
        * frame, scale, glyph, guide, facet
        * information encoding in glyphs (properties/aesthetics)
    
    #. Graph to data: What data do a I need to make my graph? [Danny]
        * **Activity:** given a plot, figure out what data would
        be necessary to create the plot
            * provide a template for the answers (x, y, shape, color, fill, etc.)
            * i.e., provide the "back story" of a plot
        * sources of plots for this activity
            * [NYT data visualizations](http://www.nytimes.com/newsgraphics/2013/12/30/year-in-interactive-storytelling/#dataviz)
            * graphics galleries
            * Danny's compilation of "science plots" from articles
            * other plots participants might know about
        * "basic plots" vs "complex plots"
            * basic: can be created by straightforward mapping from 
            a single data source
            * complex: 
                * combine elements from multiple data sources
                * plotting data is computed from the "raw data"
                    * bwplots -- `ggplot2` stats
                    * dendograms
    
    #. Data to graph
        * **Activity:** mapping data to glyphs and their properities/aesthetics [Danny]
            * provide some small  data sets (and the mapping?)
            * have participants sketch what the plot might look like.
        * data sources:
            * `NHANES`, Birds
    
#. Important data operations: [Randy]
    a. `mosaic` utilities using the formula-interface
    #. [`dplyr` toolkit](Slides/dplyrLessVolume.html)
        * `filter()`,  `mutate()`, `select()`, `summarize()`, `arrange()` 
        * `group_by()`
    #. still need to create a few "It's your turn" activities to put
    into the slides
        
    #. mention that merging data from multiple sources (joining) will be 
    discussed on Tuesday
    
    #. normalization (use Medicare spending case study) [Danny]
        * not sure what this is --rjp
        
#. Introducing `ggplot` [Randy]
    #. Data 
        * one row per glyph
        * `ggplot(data=)`
    #. Mapping aesthetics 
        * available aesthetics: x, y, colour/color, fill, shape, size, linetype,
        * `aes()`
    #. Selecting geoms and stats
        * `+ geom_()`; 
            * geom palette: line, point, bar, etc.
            * setting aesthetics
        `+ stat_()` or `+ geom_(stat=)`
            * stat palette: identity, bin, boxplot, density, function, etc
    #. Labeling
        * `+ labs()`
    #. Some short cuts
        * `qplot()`
        * `mplot()` manipulate.
    #. Bells and whistles
        * `+ theme()`
        * `+ coord_flip()`
        * changing guides
    
#. Mini-cases mini-reports

## Tuesday

1. Sources of data, and how to get it in to R
    a. Data in packages
        * `data(); `data(package=)`
        * example: `hflights::hflights`
        
    #. Data in Files
        * csv, txt -- `read.file()`
              * `read.file(file.choose())` to browse or a file
        * Excel -- `gdata::read.xls()`
        * from other software (e.g., `foreign::read.spss()`) 
      
    #. Online data [Danny]
        * Google -- `fetchGoogle()`
            * need to check that this is working 
        * Files on web sites -- `read.file()`
            * Dropbox 
            * [ons.gov.uk](http://www.ons.gov.uk/ons/datasets-and-tables)
            * [NCHS data](http://www.cdc.gov/nchs/data_access)
        #. importing data from Excel 
              * export to csv, or
              * read directly with `gdata::read.xls()`
    #. Databases [Randy]
        * [UCSC browser data](https://genome.ucsc.edu/goldenPath/help/mysql.html)
        * FAA Flights: <http://www.calvin.edu/~rpruim/talks/LessVolume/2014-05-29-GreatPlainsRUsers/dplyrLessVolume.html>
        
#. Identifying sources of data from your mini-case study

#. Specialized graphs
    a. maps
    b. networks (move to 2nd half as needed
    c. dendograms (move to 2nd half as needed
    d. interactivity (plotly, rCharts, ggvis)
    e. ask participants abot other graphs they see or would like to be
    able to create
    

#. Merging data from multiple sources (joining)
    a. **Activity:** Barbara's shiny app illustrating joins
    b. using `inner_join()`, `left_join()`, etc. (`merge()`??)
    c. strategizing data operations [Libby Shoop]
  
#. Good and bad visual displays of data
    a. Isle Royale examples [Eric Anderson]
    
#. Reports on mini-cases 

## Wednesday 

### Morning

#. More reports on mini-cases 

--------------------------

### Afternoon

1. Welcome new-comers

#. Quick review of first 2.5 days

#. How to find what you need [ ]
    a. Google
    #. CRAN task views
    #. installing packages (CRAN, github, files)
    #. documentation and vignettes
    
#. Project selection
    #. identify purpose: lectures, labs, assignments, projects, research, etc.
    #. make a list of things you want to know how to do in R
    #. beginning searching for resources (data, packages, concepts, etc.)

## Thursday


## Friday 


## Notes

The first 2.5 days includes:

* an introduction to using R for handling data and graphics
* deconstructing graphs to the underlying data
* constructing graphics from data
* data wrangling: operations to get data in the right form for graphics

The second 2.5 days could include:

* quick review of the first half of workshop
* one-on-one or small group (by discipline) development of case studies.
* summarizing data with models
* data cleaning
* data web-scraping basics 
* programming in R (writing functions)
* interactivity (Rcharts, plotly, ggvis, manipulate, shiny, etc.)

