---
title: "Wrangling Quick Start"
author: "Daniel Kaplan"
date: "Data Computing"
output: ioslides_presentation
---

```{r include=FALSE}
require(NHANES)
require(mosaic)
require(babynames)
```


## Wrangling

The process of transforming, aggregating, and combining data to put it in a glyph-ready form for your particular purpose.

### Other terms used

* Data Processing
* Data Manipulation

### *Data Computing* book uses "Wrangling" because

* "Wrangling" is distinctive and therefore rises about cognitive noise
* "Wrangling" doesn't carry misleading connotations 
    * "Processing" suggests routine
    * "Manipulation" suggests nefarious intent
    
## Five Main Kinds of "Things"

"Thing" a.k.a. "object"

1. **Data tables**: contain variables 
    Our convention: first letter is upper-case: `BabyNames`
#. **Variables**: the stuff in data tables
    Our convention: first letter is lower-case: `year`
#. **Scalars**: e.g. `"Treatment"`, `42`
#. **Functions** and their arguments: Specify actions and carry them out
    - positional arguments
    - named arguments
#. **Pipes**: Arrange sequences of actions

## Three Main Kinds of Functions for Wrangling

1. Data verbs
2. Transformation functions
3. Reduction functions

## Data Verbs

- **Always** take a data table as an input.
- May also take other arguments, e.g. 
    * variables, 
    * transformation or reduction operations
    * a second data table
- **Always** return a data table as the output

Note: Our style is **never** to use `$` or other indexing operations.  If you're using `$`, try to figure out another way.

## Transformation Functions

Always take variables and/or scalars as inputs

EXAMPLES 

A transformation expression is typically:

- `function(variable)`
- `function(variable1, variable2)`
- `function(variable, scalar)`

For instance:

- `Age > 18`
- `weight / height ^ 2`

## Reduction Functions

Take as arguments

* variable(s)
* transformation expressions

## The Data We're Going to Use

```{r}
require(babynames)
BabyNames <- babynames  # another name for the data frame
```

## Examples of Simple Wrangling Statements

```{r, results = "hide"}
filter(NHANES, Age >= 21)
select(NHANES, Age, SmokeNow)
mutate(NHANES, Age_group = ntiles(Age, 5))
summarise(NHANES, mean.age = mean(Age))
arrange(NHANES, Age)
```

Quiz: What kind of thing is each argument?

## Pipes

* Streamline the appearance of data verbs
* Support chains of operations: a step-by-step style

```{r, eval = FALSE}
NHANES %>% filter(Age >= 21)
NHANES %>% select(Age, SmokeNow)
NHANES %>% mutate(Age_group = ntiles(Age, 5))
NHANES %>% summarise(Age_group = ntiles(Age, 5))
NHANES %>% arrange(Age)
```

## Key Data Verbs

Memorize these and what they do! 

* `arrange()` -- put data in a specified order
* `filter()`  -- subset some of the rows/cases
* `select()`  -- subset some of the columns/variables
* `mutate()`  -- add a variable (column)
* `summarise()` -- summarise the data with a single row

## Working subset by subset with group_by

```{r results="hide"}
group_by(BabyNames, name, year)
BabyNames %>% group_by(name, year)
```

* `group_by()` puts the data into groups
* nothing will be apparent until you do subsequent operations

## Bringing in data from multiple sources

Also very important, but to be covered later, 

* joins such as `inner_join()`

## There are a few other things as well

Examples: `sample_n()`, `rename()`, ...

## Zebrafish

Help me edit [this file](ZebrafishRmd.html)
to show some of the data verbs 

## Try them out

* [Barbara's Shiny App](http://rstudio.calvin.edu:3838/rpruim/dataOps/)

    * This may not be working at the moment
