---
title: "Zebrafish motion"
author: "Danny Kaplan & Mary Harrington"
date: "July 14, 2015"
output: html_document
---

```{r message=FALSE,warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(printr) # for nice table printing
```

# Introduction

[Still in draft ...] Zebrafish are fish who swim ... *add more profound description*.

They are interesting *why?* [DTK: Even though I have a joint paper about the activity of Mauthner neurons in zebrafish, that probably has nothing to do with Mary's interests.

# Data as they arrived for me

Mary, this is just for our internal purposes.  It's here to document where the data came from and provide a "chain of evidence" for the validity of the data to be presented in your research publication.  (But I should have saved the original CSV file with a better name to refer to this specific run.)

```{r}
OriginalForm <- readr::read_csv("Zebrafish.csv",
                                stringsAsFactor)
OriginalForm[1:8,1:6]
```

I'm going to translate this to a long form. Think of this as cleaning more than it is wrangling.  Cleaning is often ugly.
```{r}
Tmp <- t(OriginalForm) # transpose
nms <- Tmp[1,]
Tmp <- Tmp[-1,]
nrows <- nrow(Tmp)
Tmp <- data.frame(matrix(as.numeric(Tmp), nrow = nrows),
                  stringsAsFactors=FALSE)
names(Tmp) <- nms
Activity <- gather(Tmp, "fish", "value", -time, -temp)
```

Temperature to Fahrenheit
Time to hours

Just fish A1

Just temperatures greater than 28 degrees

Mean and standard deviation of each fish

Arrange by maximum activity for each fish


