---
title: "Fetching BigQuery Data"
author: "R Pruim"
date: "July 9, 2014"
output: html_document
---

```{r, include=FALSE}
ProcessData <- FALSE
if (! ProcessData) load("Birthdays.rda")
```

This is a breif description of the process required to fetch data
from a sample data table in Google's BigQuery database.
The idea of looking here for data came from a 
[blog post](http://chmullig.com/2012/06/births-by-day-of-year/)
which pulled in births by day of year from one of the data tables.

### R Notes
The following packaes are required for the code below.
```{r, message=FALSE}
require(dplyr)
require(mosaic)
require(bigrquery)
require(magrittr)
require(lubridate)
```
The code querying Google BigQuery data needs to run in an interactive session, so
it isn't runing live while this is being knit.  Here's the session info:
```{r}
sessionInfo()
```

### Public data available at Google BigQuery

Google provides a few large data sets in their 
[sample database](https://developers.google.com/bigquery/docs/sample-tables).

Name        | Summary | # Rows
----------- | ------- | ------
gsod        | Samples from US weather stations since 1929 | 115M
mlab        | Measurement data of broadband connection performance |	240B
natality    | Birth information for the United States from 1969 to 2008 |	68M
shakespeare |	Word index for works of Shakespeare	| 164K
wikipedia	  | Revision information for Wikipedia articles	| 314M


### Code book for Natality data

The code books are availble for all of these data sets (as are
some example queries).  For example, the natality code book is available at 
[https://developers.google.com/bigquery/docs/dataset-natality](https://developers.google.com/bigquery/docs/dataset-natality).

Some notes from the code book:

* cases are individual live births.
* numerous variables are available, including year, month, day, sex, and some basic information about mother and child.
* **Important**: `record_weight` (1 or 2) indicates whether 
the data come from a region that reported 50% of births (2) or 
all births (1).  This is important if trying to count/estimate the total  number of births on a given day.

### Grabbing a Birthdays data set

#### Signing up for acces
Before you can grab data, you need to sign up for access at
[https://developers.google.com/bigquery/sign-up](https://developers.google.com/bigquery/sign-up) and create (at least one) project.  When you make your first query, you will be asked to 
authenticate, but the authentication information can be 
cached so that you only have to do that step once.

#### Fetching via an SQL query

The following code fetches the desired data from Google.
```{r eval=ProcessData}
queryString <- paste(
  "SELECT year, month, day, state, sum(record_weight) AS births",
  "FROM natality",        # name of table
  "WHERE year < 1989",    # no day-level data after this
  "group by year, month, day, state")

Birthdays.download <-
  query_exec("publicdata",     # project
             "samples",        # database
             queryString,      # query
             billing="mosaic-2014",    # project to bill to (this is free)
             max_pages=Inf)   # don't limit the number of rows
```

```{r include=FALSE, eval=FALSE}
save(Birthdays.download, file="Birthdays-download.rda")
```
```{r echo=FALSE}
if (!ProcessData) load("Birthdays-download.rda")
```


#### Post-processing with dplyr and lubridate
```{r, warning=FALSE}
Birthdays <- 
  Birthdays.download %>%
    group_by(year, month, day, state) %>% 
    summarise(births=sum(births)) %>%
    mutate(
      date = ymd(paste(year,month,day, sep="/")),
      wday = wday(date, label=TRUE, abbr=TRUE)
      ) %>%
    select(state, year, month, day, date, wday, births) %>%
  as.data.frame()
```


#### Sanity checks
There were some warnings (hidden in this output) about dates that could not be converted.
Let's see what happened.
```{r}
Birthdays %>% 
  filter(is.na(date)) %>% 
  mosaic::sample(5)
```
Those 99th days of the month look suspicious.  It isn't clear 
exactly what they code, but all of the bad dates come from these day codes that are 99.
```{r}
Birthdays %>% 
  filter(is.na(date)) %>%
  group_by(year) %>%
  summarise( n=n(), min.day=min(day), max.day=max(day))
tally( is.na(date) ~ (day==99), data=Birthdays, format="count") 
```
Let's check how the number of bad codes varies by state
```{r}
Birthdays %>% 
  group_by(state) %>%
  summarise(n=n(), bad=sum(day > 50)) %>%
  arrange(bad/n)
```  
Let's filter those 99's out
```{r}
Birthdays <-
  Birthdays %>% filter(day < 99)
```
```{r eval=ProcessData}
#### Saving in R native format for later use.
save(Birthdays, file="Birthdays.rda")
```