
---
title: "dplyr presentation"
author: "Jo Hardin"
date: "June 21, 2016"
output: 
  ioslides_presentation: 
    smaller: yes
    top: 50%
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, size="small",
                      fig.width=8, fig.height=6)
```

## Tidy Data 

- rows (cases/observational units) and 
- columns (variables).  

- The key is that *every* row is a case and *every* column is a variable.  

- No exceptions.

## Chaining 

The pipe syntax (%>%) takes a data frame (or data table) and sends it to the argument of a function.  


- x %>% f(y) is the same as f(x, y)

- y %>% f(x, ., z) is the same as f(x,y,z)


## Building Tidy Data 

- object_name <- function_name(arguments) 
- object_name <- data_table %>% function_name(arguments) 
- object_name <-   
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;data_table %>%  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; function_name(arguments) %>%  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; function_name(arguments)

- in chaining, the value (on left) %>% is **first argument** to the function (on right)


## 5 Main Data Verbs

Data verbs take data tables as input and give data tables as output

1. select(): removes unwanted *variables* (and rename() )
2. filter():  removes/subsets unwanted *cases*
3. mutate(): transforms the variable (and transmute() like mutate, returns only new variables)
4. arrange(): reorders the cases
5. summarize(): computes summary statistics

## Other Data Verbs
- distinct(): returns the unique values in a table
- sample_n(): take a random row(s)
- head():  grab the first few rows
- tail(): grab the last few rows
- group_by(): SUCCESSIVE functions are applied to groups
- ungroup():  reverse the grouping action
- summarise():  
    + min(), max(), mean(), sum(), sd(), median(), and IQR()
    + n(): number of observations in the current group
    + n_distinct(): number of unique values
    + first_value(), last_value(), and nth_value(x, n): (like x[1], x[length(x)], and x[n] )

## You should know that I did this:

```{r message=FALSE, warning=FALSE, echo=TRUE}
require(mosaic)
require(babynames)
require(NHANES)
Babynames = babynames
names(NHANES) <- tolower(names(NHANES))
```

## Finally, some Examples! 

```{r}
Babynames %>% nrow()
Babynames %>% names()
```

## Finally, some Examples! 

```{r}
Babynames %>% glimpse()
```

## Finally, some Examples!

```{r}
Babynames %>% head()
```

## Finally, some Examples! 

```{r}
Babynames %>% sample_n(size=5)
```


## NHANES Data

```{r}
names(NHANES)
```

## 1. select(): removes unwanted *variables*


find the sleep variables

```{r}
NHANESsleep <- 
  NHANES %>% 
  select(gender, age, weight, race1, race3, education, sleeptrouble,
         sleephrsnight, tvhrsday, tvhrsdaychild, physactive)

names(NHANESsleep)
dim(NHANESsleep)
```

## 2. filter():  removes/subsets unwanted *cases* 

subset for college students

```{r fig.width=5, fig.height=3, fig.center=TRUE}
NHANESsleep <- NHANESsleep %>% filter(age %in% c(18:22))
histogram(~age, data=NHANESsleep)
```

## 3. mutate(): transforms the variable 

mutate or transmute to create a new variable?

```{r }
NHANESsleep %>% mutate(weightlb = weight*2.2) %>% head(3)
```

## 3. mutate(): transforms the variable 

mutate or transmute to create a new variable?

```{r }
NHANESsleep %>% transmute(weightlb = weight*2.2) %>% head(3)
```

## 5. summarize(): computes summary statistics

```{r}
# number of people (cases) in NHANES
NHANES %>% summarise(n())
```

## 5. summarize(): computes summary statistics

```{r}
# total weight of all the people in NHANES (silly)
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  summarise(sum(weightlb, na.rm=TRUE))
```

## 5. summarize(): computes summary statistics

```{r}
# mean weight of all the people in NHANES
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  summarise(mean(weightlb, na.rm=TRUE))
```

## 5. summarize() with group_by()

```{r}
# mean weight of all the people in NHANES, broken down by education
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  group_by(education) %>% 
  summarise(mean(weightlb, na.rm=TRUE))
```

## 4. arrange(): reorders the cases

```{r}
# mean weight, by education, sorted
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  group_by(education) %>% 
  summarise(avewt = mean(weightlb, na.rm=TRUE)) %>% 
  arrange(avewt)
```

## Your Turn 

When starting, it can be helpful to work with a small subset of the data.  When you have your data wrangling statements in working order, shift to the entire data table.

```{r}
SmallSubset <-
  Babynames %>%
  filter(year > 2000) %>%
  sample_n(size = 200)

names(SmallSubset)
```


## How many babies are represented?

```{r eval=FALSE}
SmallSubset %>%
  summarise(total = ????(n)) # a reduction verb
```

## How many babies are represented?

```{r}
SmallSubset %>%
  summarise(total=sum(n))
```


## How many babies are there in each year?

```{r eval=FALSE}
SmallSubset %>% 
  group_by(????) %>% 
  summarise(total = ????(n))
```

## How many babies are there in each year?

```{r}
SmallSubset %>% 
  group_by(year) %>% 
  summarise(total = sum(n))
```

## How many distinct names in each year?

```{r eval=FALSE}
SmallSubset %>%
  group_by(????) %>%
  summarise(name_count = n_distinct(????))
```

## How many distinct names in each year?

```{r}
SmallSubset %>%
  group_by(year) %>%
  summarise(name_count = n_distinct(name))
```

## How many distinct names of each sex in each year?

```{r eval=FALSE}
SmallSubset %>%
  group_by(????, ????) %>%
  summarise(????)
```

## How many distinct names of each sex in each year?

```{r}
SmallSubset %>%
  group_by(year, sex) %>%
  summarise(name_count=n_distinct(name))
```


## Track the yearly number of Janes and Marys over the years.

```{r eval=FALSE}
Result <-
  Babynames %>%
  ????(name %in% c("Jane", "Mary")) %>% # just the Janes and Marys
  group_by(????, ????) %>% # for each year for each name
  summarise(count = ????)
```

## Track the yearly number of Janes and Marys over the years.

```{r}
Result <-
  Babynames %>%
  filter(name %in% c("Jane", "Mary")) %>% # just the Janes and Marys
  group_by(name, year) %>% # for each year for each name
  summarise(count = sum(n))
```

## Plot out the result

Put `year` on the x-axis and the count of each name on the y-axis.  Note that `ggplot()` commands use `+` rather than `%>%`.

```{r eval=FALSE}
ggplot(data=Result, aes(x = year, y = count)) +
  geom_point()
```

* *Map* the name (Mary or Jane) to the aesthetic of color.  Remember that mapping to aesthetics is always done inside the `aes()` function.
* Instead of using dots as the glyph, use a line that connects consecutive values: `geom_line()`.
* Change the y-axis label to "Yearly Births": `+ ylab("Yearly Births")`
* *Set* the line thickness to `size=2`. Remember that "setting" refers to adjusting the value of an aesthetic to a constant.  Thus, it's *outside* the `aes()` function.

## Plot out the result

```{r fig.width=5, fig.height=3, fig.center=TRUE}
ggplot(data=Result, aes(x = year, y = count)) +
  geom_line(aes(color=name), size=2) + 
  ylab("Yearly Births")
```


## Look at the *proportion* of births rather than the count

```{r eval=FALSE}
Result2 <-
  Babynames %>%
  group_by(year) %>%
  mutate(total = ????(n)) %>%
  filter(????) %>%
  mutate(proportion = n / total)
```


## Look at the *proportion* of births rather than the count

Just as you did with count vs year, graph proportion vs year.
```{r eval=FALSE}
Result2 %>%
  Your ggplot statements go here!
```
* Add a vertical line to mark a year in which something happened that might relate to the increase or decrease the popularity of the name.  Example: The movie [*Whatever Happened to Baby Jane*](http://en.wikipedia.org/wiki/What_Ever_Happened_to_Baby_Jane%3F_%281962_film%29) came out in 1962.  The glyph is a vertical line: `geom_vline()`.

## Look at the *proportion* of births rather than the count

```{r}
Result2 <-
  Babynames %>%
  filter(sex == "F") %>%
  group_by(year) %>%
  mutate(total = sum(n)) %>%
  filter(name %in% c("Mary", "Jane")) %>%
  mutate(proportion = n / total)
```

## Look at the *proportion* of births rather than the count

```{r fig.width=5, fig.height=3, fig.center=TRUE}
ggplot(data=Result2, aes(x=year, y=proportion)) +
  geom_line(aes(color=name), size=2) +
  ylab("Yearly Births") +
  geom_vline(xintercept=1962)
```

 
  
##  Pick out name(s) of interest to you 

Plot out their popularity over time.

##  Pick out name(s) of interest to you 

```{r fig.width=5, fig.height=3, fig.center=TRUE}
Frances <- babynames %>%
  filter(name== "Frances") 
```

##  Pick out name(s) of interest to you 

```{r fig.width=5, fig.height=3, fig.center=TRUE}
Frances %>% ggplot(aes(x=year, y=n)) +
  geom_point(aes(color=sex), size=.5) + 
  geom_line(aes(color=sex)) + 
  geom_vline(xintercept=2006) + scale_y_log10() +
  ylab("Yearly total on log10 scale")
```


