---
title: "Some dplyr tricks"
author: "R Pruim"
date: "June 25, 2016"
output: 
  html_document: 
    fig_height: 2.5
    fig_width: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, warning = FALSE}
require(mosaic)
require(lazyeval)
theme_set(theme_minimal())
```

Nick asked how to plot 1-sample t confidence intervals for subsets defined be each level 
of a factor.  Here are two ways to do it.  Both methods 

  * use `confint()` to extract just the interval from the output of `t.test()`
  * use `lapply()` to create a list of these results -- 1 per level of the factor
  * use `bind_rows()` to combine this list into a single data frame.
  * use `mutate()` to add on a variable indicating which test is from which level

The difference between the two methods is in the filtering down to the subset of
values matching just one level.

### Method 1: use interp() to mix inside- and outside- the data values

```{r}
D1 <- 
  bind_rows(
    lapply( 
      levels(HELPrct$substance),
      function(l) { 
        confint(t.test( ~ age, data = HELPrct %>% filter_(interp( ~ substance == g, g = l))))
      }
    )
  ) %>% 
  mutate ( level = levels(HELPrct$substance) )
```

### Method 2: Use a scratch variable

```{r}
D2 <- 
  bind_rows(
    lapply( 
      levels(HELPrct$substance),
      function(l) { 
        confint(
          t.test( ~ age, 
                  data = HELPrct %>% mutate(target = l) %>%
                    filter(substance == target)
          ))
      }
    )
  ) %>% 
  mutate ( level = levels(HELPrct$substance) )

identical(D1, D2)
```


### Plotting is the same either way

```{r}
ggplot(data = D2) +
  geom_pointrange( 
    aes(x = level, y = `mean of x`, ymin = lower, ymax = upper) ) +
  coord_flip()
```
