---
title: "An exercise on data operations"
author: "DTK"
date: "July 10, 2014"
output: html_document
---

```{r include=FALSE}
require(DCFdevel)
require(dplyr)
require(mosaic)
```

## Broad Description

Your task is to find the fraction of the population in each country that lives in cities larger than 100,000.

## The Data

For the purposes of this exercise, you will use **three** data frames:
```{r}
require(DCFdevel)
require(mosaic)
data(WorldCities) # Creates WorldCities
TotalPop <- CIAdata(2119) # population data
CountryCodes <- fetchData("countrycodes.csv")
```

Each of these frames has information about the identity of a country, but perhaps in different forms.  By grouping, summarizing, joining, and arithmetic, you should be able to construct the answer to the problem.

ISO2 and ISO3 are different, but related, official standards for identifying countries.

## Tools

*The `dplyr()` functions such as `group_by()`, `mutate()`, `summarize()`, etc. 
*The `DCFdevel::toISO3()`.  See the documentation for that function.  Here's an example:
```{r}
toISO3(c('Canada','England','Great Britain', 'United Kingdom'))
```

## Process

Before writing the computer commands to perform the operation:

* Look at the names and contents of the variables in the different data frames.
* Draw a flow diagram of what processes will be required.  
    Only then, translate your process into computer instructions.
    
    
    
## A Solution

Here are the input files (with just the relevant variables):

```{r include=FALSE}
set.seed(100)
```

#### TotalPop

Each case is a country
```{r echo=FALSE}
select(TotalPop, country, pop) %>% sample_n(3) %>% 
  print(row.names=FALSE)
```

The country is identified by a conventional name, spelled out in full.  This can be converted to an ISO code using `toISO3()`.

#### WorldCities

Each case is a city:
```{r echo=FALSE}
select(WorldCities, Country, Population)  %>% sample_n(3) %>% 
  print(row.names=FALSE)
```

The country is identified by a 2-letter ISO code, such as BR or IN

#### CountryCodes

Each case is a country:
```{r echo=FALSE}
select(CountryCodes, iso3, iso2)  %>% sample_n(3) %>% 
  print(row.names=FALSE)
```

General plan:

* 


* Create a data table, `UrbanPopByCountry` which has the urban population for each country

* Combine `UrbanPopByCountry` with the `TotalPopByCountry` country-wide population for each county.  This will look like 
    
    Country | UrbanPop | Total Pop
    --------|----------|-----------
      BEL   |  7221658 | 10449361
      ...   |  ...     | ....