---
title: "An exercise on data operations"
author: "DTK"
date: "July 10, 2014"
output: html_document
---

```{r include=FALSE}
require(DCFdevel)
require(mosaic)
```

## Broad Description

Your task is to find the fraction of the population in each country that lives in cities larger than 100,000.

## The Data

For the purposes of this exercise, you will use **three** data frames:
```{r}
require(DCFdevel)
require(mosaic)
data(WorldCities) # Creates WorldCities
pop <- CIAdata(2119) # population data
codes <- fetchData("countrycodes.csv")
```

Each of these frames has information about the identity of a country, but perhaps in different forms.  By grouping, summarizing, joining, and arithmetic, you should be able to construct the answer to the problem.

ISO2 and ISO3 are different, but related, official standards for identifying countries.

## Tools

*The `dplyr()` functions such as `group_by()`, `mutate()`, `summarize()`, etc. 
*The `DCFdevel::toISO3()`.  See the documentation for that function.  Here's an example:
```{r}
toISO3(c('Canada','England','Great Britain', 'United Kingdom'))
```

## Process

Before writing the computer commands to perform the operation:

* Look at the names and contents of the variables in the different data frames.
* Draw a flow diagram of what processes will be required.  
    Only then, translate your process into computer instructions.