---
title: "Joins"
author: "Danny Kaplan"
date: "CVC 2015"
output: ioslides_presentation
---

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(printr)
```

## Combining Two Tables

All of the data verbs we have used up to now take a single data table as an input.

Sometimes you need to combine data between different sources.

* Perhaps they come from different experiments or institutions.
* Perhaps they are completely different types of data, e.g. medical treatment records and demographic or education data.
    - Example: `MedicareCharges`, `MedicareProviders`, `DirectRecoveryGroups`.
    - Example: the registrar's database

## Row-wise concatenation

Slapping one data table onto another row-wise is *sometimes* an appropriate activity, but it is ... **not a join!**

Example of using row-wise concatenation:

* Each day's new earthquakes are available through a geophysics site.  You're collecting all the events over a long period of time.

## Column-wise concatenation

You're TA just entered the students grades for assignment 7 into a spreadsheet, sorted by student name.

You have your master spreadsheet with assignments 1 through 6.

Concatenate the assignment-7 grades as a new variable for your master spreadsheet.


** Watch out!** This is a brittle process. 

## Things that can go wrong here

* Suppose your TA's sheet has sorted the students by last name and yours has them sorted by first name.
* Suppose you delete students who drop the course, but the TA doesn't have this information.

## Instead ...

1. Match the students in one table to those in the other.
2. Gently lift the Assignment-7 data for each student from the TA's table into the appropriate cell in your table.

This is one example of a join. There are other kinds of join for situations where there might be multiple matches between cases in the two tables.

## Example: Zebrafish

Continuation of the example [here](../Wrangling/Zebrafish.html) 


## Activity: Grade-point averages

Here are three data tables relating student grades in courses at Macalester in 2005 (?)

```{r}
grades <- read.csv("http://tiny.cc/mosaic/grades.csv",
                   stringsAsFactors = FALSE)
courses <- read.csv("http://tiny.cc/mosaic/courses.csv",
                    stringsAsFactors = FALSE)
grade_to_number <- read.csv("http://tiny.cc/mosaic/grade-to-number.csv",
                            stringsAsFactors = FALSE)
```

## Activity: Tasks

1. How many students in each department?
#. What fraction of grades are below B+ in each department?
#. What's the grade-point average (GPA) for each student?
#. What's the 95% confidence interval on the GPA for each student?
#. Grade-point average for each department or instructor
#. (Statistically more sophisticated) To what extent does the grade reflect the student or the department or instructor?

## Joining for cleaning

Birds at the Ordway Nature Preserve

```{r eval=FALSE}
devtools::install_github("DataComputing/DataComputing")
require(DataComputing)
View(OrdwayBirds)
help(OrdwayBirds)
```

As a group, let's correct the species names using [this data table](https://docs.google.com/spreadsheets/d/1GDKeulVSRXL7Q_5Q0a7Offc_IPX2T4sZ80VN5VLASew/edit#gid=0).

## Task for the birds

Count how many birds there are of each species in `OrdwayBirds` using the corrected species names.


