---
title: "Everything you've ever wanted to know about t"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggformula)
library(mosaic)
set.seed(103)
knitr::opts_chunk$set(echo = FALSE)
make_data <- function(n, delta_means=10){
  set.seed(103)
  n_group_a <- n/2
  n_group_b <- n - n_group_a
  res <- data_frame(
    vals = c(rnorm(n_group_a, mean = 100, sd = 10), 
             rnorm(n_group_b, mean = 100 + delta_means, sd = 15)), 
    group = c(rep("Sharks", n_group_a), rep("Jets", n_group_b))
  )
  set.seed(NULL)
  
  res
}
```

## Assumptions about the t test

* It is 1908 and we haven't learned any more modern method.
* We're interested in the difference between two groups, and there's nothing else in the world that might influence the response variable than the grouping.

### Making up data

We'll have two groups, one will have a mean of 100 and the other of 110. We'll have $n$ cases altogether, evenly split.

```{r echo = TRUE}
n <- 20
n_group_a <- n/2
n_group_b <- n - n_group_a
Made_up <- data_frame(
  vals = c(rnorm(n_group_a, mean = 100, sd = 10), 
            rnorm(n_group_b, mean = 110, sd = 15)), 
  group = c(rep("Sharks", n_group_a), rep("Jets", n_group_b))
  )
```

```{r}
Made_up <- make_data(n = 20)
Made_up %>%
  gf_boxplot(vals ~ group) %>%
  gf_point(vals ~ group, alpha = 0.25) %>%
  gf_refine(coord_flip())
```

# The t test

```{r}
make_data(n=20) %>%
  t.test(vals ~ group, data = .)
```

Some commentary

### Exercise 

Do what I say ...


```{r firstpass, exercise = TRUE}
make_data(n=20) %>%
  t.test(___ ~ ___, data = .)
```

```{r firstpass-solution}
make_data(n=20) %>%
  t.test(vals ~ group, data = .)
```



<div id="firstpass-hint">What ever you want for your hint. And you can put as much here as you care to.

* Bullet one
* The **second** point
* $\pi$
</div>

```{r Q1}
question("Does p get smaller or larger as n increases?",
         answer("large n tends to give large p", correct=FALSE,
                message = "Did you read the book this morning?"),
         answer("small n tends to give large p", correct = TRUE),
         answer("It makes not a whit of difference", correct = FALSE,
                message = "Don't bring your **nihilism** to my class! $e^{i \\pi} = -1$"),
        allow_retry = TRUE,
        random_answer_order = TRUE)
```